---
output:
  github_document:
    pandoc_args: ["--wrap=preserve"]
---

```{r setup, include=FALSE}
library(duckknit)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::knit_engines$set(sql = knitr::knit_engines$get("duckdb"))
knitr::knit_hooks$set(document = function(x) gsub("(?m)^``` duckdb", "```sql", x, perl = TRUE))
cli <- Sys.getenv("DUCKDB_CLI")
stopifnot(nzchar(cli), file.exists(cli))
options(duckknit.duckdb = cli)
local({
  ns <- asNamespace("duckknit")
  proc <- processx::process$new(cli, args = c("-unsigned", ":memory:"),
    stdin = "|", stdout = "|", stderr = "|", cleanup = TRUE, cleanup_tree = TRUE)
  unlockBinding(".sessions", ns)
  ns$.sessions[["readme"]] <- proc
  lockBinding(".sessions", ns)
  unlockBinding(".state", ns)
  ns$.state$last_session <- "readme"
  lockBinding(".state", ns)
  extension <- normalizePath("build/release/duckvep.duckdb_extension", mustWork = TRUE)
  load <- duckknit:::duckknit_exec(proc, sprintf("LOAD '%s';", extension))
  if (nzchar(trimws(load$stderr))) stop(load$stderr)
  duckknit:::duckknit_exec(proc, "SET lambda_syntax='ENABLE_SINGLE_ARROW';")
  fixture <- paste(readLines("test/data/duckvep/readme.sql"), collapse = "\n")
  setup <- duckknit:::duckknit_exec(proc, fixture)
  if (nzchar(trimws(setup$stderr))) stop(setup$stderr)
})
fmt <- function(x) format(x, big.mark = ",", scientific = FALSE, trim = TRUE)
history <- read.csv("test/duckvep/conformance/data/conformance_history.csv")
scope <- subset(history, engine == "duckvep" & stratum_kind == "all" &
  model %in% c("final-dbsnp", "final-grch37", "plasmodium-falciparum-63"))
scope <- scope[order(scope$run_date, scope$source_revision), ]
scope <- scope[!duplicated(scope$model, fromLast = TRUE), ]
hgvs <- read.csv("benchmarks/data/duckvep_fastvep/conformance.csv")
hgvs <- subset(hgvs, engine == "duckvep")
metric <- function(name) hgvs$n[match(name, hgvs$metric)]
throughput <- read.csv("benchmarks/data/duckvep_throughput.csv")
throughput <- subset(throughput, workload == "ensembl116_grch38_giab_sites_hash40" &
  output_mode == "rich" & threads == 1)
throughput <- throughput[order(throughput$run_date), ]
```

# DuckVEP

DuckVEP implements Ensembl VEP 116 consequence semantics as a resident, relation-first DuckDB kernel. Models are loaded from ordered relations; annotation, transcript HGVS, and phased haplotypes remain queryable relations. The extension uses htslib for indexed reference access, not a DuckHTS runtime dependency.

## Evidence

```{r evidence, results='asis', echo=FALSE}
cat("| Evidence | Pairs / variants | Exact | Different | Unresolved | Variants/s | Receipt |\n|:--|--:|--:|--:|--:|--:|:--|\n")
for (i in seq_len(nrow(scope))) {
  row <- scope[i, ]
  cat(sprintf("| %s | %s | %s | %s | %s | — | [conformance history](test/duckvep/conformance/data/conformance_history.csv) |\n",
    row$model, fmt(row$n), fmt(row$exact_agree),
    fmt(row$n - row$exact_agree - row$unresolved), fmt(row$unresolved)))
}
for (part in c("hgvsc", "hgvsp")) {
  present <- metric("shared_keys") - metric(paste0(part, "_suffix_both_absent"))
  exact <- metric(paste0(part, "_suffix_exact_present"))
  discordant <- metric(paste0(part, "_suffix_discordant"))
  cat(sprintf("| GRCh38 %s suffix agreement (present) | %s | %s | %s | %s | — | [executable-VEP comparison](benchmarks/data/duckvep_fastvep/conformance.csv) |\n",
    if (identical(part, "hgvsc")) "HGVSc" else "HGVSp",
    fmt(present), fmt(exact), fmt(discordant), fmt(present - exact - discordant)))
}
row <- tail(throughput, 1L)
cat(sprintf("| GRCh38 sorted rich-output, single thread | %s | — | — | — | %s | [timing receipt](benchmarks/data/duckvep_throughput.csv) |\n",
  fmt(row$variants), fmt(round(row$variants_per_second))))
```

Each corpus result applies to the named receipt and model, not an arbitrary deployment sample. See the [conformance report](benchmarks/duckvep_conformance.md) and [throughput report](benchmarks/duckvep_throughput.md) for methods and limitations.

## Query a model

The [small model setup](test/data/duckvep/readme.sql) loads a transcript and exons and stages a variant from a [committed Parquet fixture](test/data/duckvep/minimal_bcsq.parquet). The README runs against a locally built extension in a DuckDB CLI session.

```{sql annotate, eval=TRUE}
SELECT event_index, consequence_mask, transcript_hgvs, transcript_hgvs_status
FROM duckvep_annotate('readme_events', 'readme', hgvs := true);
```

Calls are relations too:

```{sql haplotypes, eval=TRUE}
CREATE TABLE readme_calls AS SELECT event_index, seq_region, position,
  reference, alternate, 1::UINTEGER alt_index, 0::UINTEGER transcript_index,
  0::UBIGINT sample_index, [1,0]::INTEGER[] alleles,
  [false,true]::BOOLEAN[] phase_before, NULL::BIGINT phase_set
FROM readme_events;
SELECT transcript_index, carrier_count, sequence_status,
  length(contributors) AS contributors
FROM duckvep_haplotypes('SELECT * FROM readme_calls', 'readme');
```

The [design](design/duckvep.md) describes model and consequence contracts. [Rduckvep](r/Rduckvep) builds the same sources offline and provides a connection and haplotype front end; DuckHTS is optional for VCF reading.
