---
output:
  github_document:
    pandoc_args: ["--wrap=preserve"]
---

<!-- README.md is generated from README.Rmd using duckknit. Please edit that file. -->

```{r setup, include=FALSE}
library(duckknit)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::knit_engines$set(sql = knitr::knit_engines$get("duckdb"))
knitr::knit_hooks$set(document = function(x) gsub("(?m)^``` duckdb", "```sql", x, perl = TRUE))
cli <- Sys.getenv("DUCKDB_CLI")
stopifnot(nzchar(cli), file.exists(cli))
options(duckknit.duckdb = cli)
local({
  ns <- asNamespace("duckknit")
  proc <- processx::process$new(cli, args = c("-unsigned", ":memory:"),
    stdin = "|", stdout = "|", stderr = "|", cleanup = TRUE, cleanup_tree = TRUE)
  unlockBinding(".sessions", ns)
  ns$.sessions[["readme"]] <- proc
  lockBinding(".sessions", ns)
  unlockBinding(".state", ns)
  ns$.state$last_session <- "readme"
  lockBinding(".state", ns)
  extension <- normalizePath("build/release/duckvep.duckdb_extension", mustWork = TRUE)
  load <- duckknit:::duckknit_exec(proc, sprintf("LOAD '%s';", extension))
  if (nzchar(trimws(load$stderr))) stop(load$stderr)
  fixture <- paste(readLines("test/data/duckvep/readme.sql"), collapse = "\n")
  setup <- duckknit:::duckknit_exec(proc, fixture)
  if (nzchar(trimws(setup$stderr))) stop(setup$stderr)
})
source("scripts/readme-evidence.R")
evidence <- readme_evidence()
n <- function(x) format(x, big.mark = ",", scientific = FALSE, trim = TRUE)
conf <- evidence$conformance
hgvs <- evidence$hgvs
haplotype <- evidence$haplotype
pub <- evidence$public
```

# DuckVEP

[![Extension CI](https://github.com/RGenomicsETL/DuckVEP/actions/workflows/MainDistributionPipeline.yml/badge.svg)](https://github.com/RGenomicsETL/DuckVEP/actions/workflows/MainDistributionPipeline.yml)
[![Site](https://github.com/RGenomicsETL/DuckVEP/actions/workflows/pages.yml/badge.svg)](https://rgenomicsetl.github.io/DuckVEP/)

**Variant effect prediction as a SQL relation.** Load an Ensembl transcript model into DuckDB once, then query `duckvep_annotate_sql(...)` as a relation to filter, join and aggregate alongside VCF, ClinVar or population-frequency data. A new question is a new query, not another pipeline run.

DuckVEP returns typed Sequence Ontology consequences and impact, cDNA/CDS/protein positions and amino acids, NMD predictions, and HGVS. Phased calls also support coding whole-haplotype consequences.

Independent-event consequences and HGVS are compared with executable Ensembl VEP `r sub("^VEP ", "", conf$oracle[[1L]])` and its matching core and variation databases. Discordant and unresolved pairs stay in the counts, deliberate differences are documented, and unavailable results carry a status and reason.

## Try it

A model is loaded from ordinary relations: regions, transcripts with their CDS and flanking sequence, and exons. In production these come from an Ensembl core database through `query(duckvep_ensembl_regions_sql(...))` and `query(duckvep_ensembl_transcripts_sql(...))`. Here a [one-transcript fixture](test/data/duckvep/readme.sql) stands in:

```{sql model, eval=TRUE}
SELECT loaded FROM duckvep_model_load('demo',
  'SELECT * FROM readme_regions ORDER BY seq_region',
  'SELECT * FROM readme_transcripts ORDER BY seq_region, transcript_start',
  'SELECT * FROM readme_exons ORDER BY transcript_index, exon_start');
```

The named model stays resident in the DuckDB session. Annotate seven variants across the transcript, from the 5′ UTR to the intron, and join the results back to their input in one query:

```{sql annotate, eval=TRUE}
CREATE TABLE demo_events AS
SELECT row_number() OVER (ORDER BY position, alternate)::UBIGINT AS event_index,
       1::UINTEGER AS seq_region, position::UBIGINT AS position, reference, alternate,
       NULL::UBIGINT AS end_position, NULL::VARCHAR AS structural_type,
       NULL::VARCHAR AS copy_change, NULL::UINTEGER AS mate_seq_region,
       NULL::UBIGINT AS mate_position
FROM (VALUES (110, 'C', 'T'), (124, 'T', 'C'), (125, 'A', 'G'), (130, 'C', 'CA'),
             (134, 'C', 'A'), (151, 'G', 'A'), (175, 'A', 'G')) v(position, reference, alternate);

SELECT e.position, e.reference || '>' || e.alternate AS change,
       a.consequence, a.impact, a.transcript_hgvs, a.protein_hgvs, a.nmd_prediction
FROM query(duckvep_annotate_sql('demo_events', 'demo', struct_pack(rich := true, hgvs := true))) a
JOIN demo_events e USING (event_index)
ORDER BY e.position;
```

The frameshift and the stop gain escape nonsense-mediated decay because they fall in the first CDS positions (`nmd_escape_early_cds`).

Some HGVS fields are empty, and that isn't a parsing failure. This model was loaded without a reference FASTA, and those notations need flanking sequence to shift and render. DuckVEP says so rather than guess, and asking why is just another query:

```{sql unresolved, eval=TRUE}
SELECT e.position, a.transcript_hgvs_status, a.transcript_hgvs_reason
FROM query(duckvep_annotate_sql('demo_events', 'demo', struct_pack(hgvs := true))) a
JOIN demo_events e USING (event_index)
WHERE a.transcript_hgvs IS NULL
ORDER BY e.position;
```

Phased genotypes become whole-haplotype consequences, again as a relation. The changes at 124 and 125 fall in the same codon (Val2). On their own, one is missense and the other synonymous. Carried on the same haplotype, they are read together as one codon, GTA to GCG, and classified once:

```{sql haplotypes, eval=TRUE}
CREATE TABLE demo_calls AS
SELECT event_index, seq_region, position, reference, alternate,
       1::UINTEGER alt_index, 0::UINTEGER transcript_index, 0::UBIGINT sample_index,
       [1, 0]::INTEGER[] alleles, [false, true]::BOOLEAN[] phase_before, NULL::BIGINT phase_set
FROM demo_events WHERE position IN (124, 125);

SELECT carrier_count, length(contributors) AS contributors, prediction_status,
       haplotype_consequences, haplotype_impact
FROM duckvep_haplotypes('SELECT * FROM demo_calls', 'demo');
```

For VCF or BCF input, [`duckvep_coding_calls(model, path)`](docs/functions.md#duckvep_coding_calls) builds the calls relation directly; its model's `seq_region_name` values must match VCF `CHROM`. It discovers coding transcripts before decoding GT and PS, then feeds `duckvep_haplotypes`. If you stage calls yourself, `duckvep_coding_transcripts(model, seq_region, position, reference, alternate)` returns the transcripts touched by the normalized coding event, so a shared indel anchor alone does not count. R wrappers are `Rduckvep::rduckvep_coding_calls()` and `Rduckvep::rduckvep_coding_transcripts()`.

## How close to VEP

Across `r n(nrow(conf))` recorded corpora, executable VEP 116 comparisons contain **`r n(sum(conf$exact))` exact of `r n(sum(conf$pairs))` transcript and feature pairs**, with `r n(sum(conf$unresolved))` unresolved and `r n(sum(conf$differences))` discordant pairs. The corpora cover GRCh37 and GRCh38, *Plasmodium falciparum* genetic codes, small variants, exact structural events, paired breakends, regulatory features and motifs. The corpus release criterion is zero discordant, missing, extra or unresolved pairs.

The same evidence set records `r n(hgvs$exact[hgvs$field == "HGVSc"])` exact HGVSc and `r n(hgvs$exact[hgvs$field == "HGVSp"])` exact HGVSp transcript pairs against VEP `--hgvs`; `r n(hgvs$clinvar_pairs[[1L]])` pairs come from ClinVar chromosome 21, with `r n(sum(hgvs$discordant))` discordant. Methods and term-level results are in the [conformance report](benchmarks/duckvep_conformance.md); the [consequence](test/duckvep/conformance/data/conformance_history.csv) and [HGVS](test/duckvep/conformance/data/hgvs_history.csv) ledgers hold per-run counts and source revisions. Ledger revisions preceding extraction from DuckHTS resolve through the [commit map](design/duckhts-commit-map.txt).

<details>
<summary>Per-corpus consequence counts</summary>

```{r conformance, echo=FALSE, results='asis'}
shown <- conf
shown$pairs <- n(shown$pairs)
shown$exact <- n(shown$exact)
shown$unresolved <- n(shown$unresolved)
shown$differences <- n(shown$differences)
knitr::kable(shown, col.names = c("Corpus", "Assembly", "Oracle", "Pairs", "Exact", "Unresolved", "Different"),
  align = c("l", "l", "l", "r", "r", "r", "r"), format = "pipe")
```
</details>

## Whole haplotypes

VEP 116 does not define whole-haplotype consequences. The coding-only [`duckvep-coding` contract](design/duckvep_haplotype_contract.md) uses pinned `bcftools csq` where transcript models and phase semantics are comparable. Independent base-R goldens and hand-derived checks cover supported cases outside csq's domain: non-diploid calls, alleles over 50 bases, nonstandard genetic codes, and incomplete CDS starts or ends. Unsupported inputs retain explicit statuses and reasons.

The pinned full-HG002 qualification for [issue #34](https://github.com/RGenomicsETL/DuckVEP/issues/34) recorded one-core, cold-process medians against the Ensembl 116 model: `r sprintf("%.2f", haplotype$csq_s)` s for csq, `r sprintf("%.2f", haplotype$cli_s)` s for the fused-reader CLI path (`r sprintf("%.2f", haplotype$cli_speedup)`×, including DuckVEP model load), and `r sprintf("%.2f", haplotype$r_s)` s for the R path (`r sprintf("%.2f", haplotype$r_speedup)`×). The [qualification method and receipts](benchmarks/data/haplotype_scale/README.md) identify the input and run contract.

## Where it differs, and why

Compatibility is claimed only where it has been measured, and differences are kept, not hidden. The [compatibility and errata record](ERRATA.md) separates three verdicts, each with its witness:
- a VEP convention DuckVEP must follow;
- a gap in DuckVEP;
- a likely upstream erratum.

The main cases:

- **Published release annotations are not the oracle.** Ensembl's release VCFs sometimes disagree with executable VEP. In release 116, `X/Y:276322 G>A` is published as `intergenic_variant`, while the executable emits three `5_prime_UTR_variant` rows per chromosome. DuckVEP follows the executable, and the [PAR witnesses](test/duckvep/conformance/README.md) pin both.
- **Breakends are evaluated one event at a time.** VEP 116's buffered breakend path uses a chromosome-blind interval tree and can drop valid transcript pairs in multi-chromosome batches. The oracle runs with `--buffer_size 1` to isolate that; DuckVEP has no such batching effect ([paired-breakend differential](benchmarks/duckvep_conformance.md#paired-breakend-differential)).
- **Imprecise structural variants use the nominal span**, as VEP's registered predicates do. `CIPOS` and `CIEND` stay on the row as metadata rather than being dropped.
- **Structural-span comparison:** VEP skips events above `--max_sv_size` (5 kb by default). For supported exact structural events, DuckVEP uses the nominal span without that cutoff; conformance comparisons set VEP's limit to 10 Mb.
- **Circular regions use lifted intervals.** Origin-crossing transcripts, exons and regulatory or motif features preserve overlap and HGVS behavior across the origin. This is separate from mitochondrial translation; human MT has no wrapped object. VEP cannot oracle origin-crossing models, so they are checked by rotation equivariance and linear-model differential ([design](design/duckvep.md), [survey](benchmarks/data/circular_source_survey.md)).
- **Unknown is a value.** When a result can't be computed, the row carries `duckvep_status` and a reason instead of a guess. Examples are a missing reference sequence, a reference mismatch, or an unsupported symbolic allele.
- **Outside the claim:**
  - species and releases that haven't been tested;
  - compound (cis) HGVS presentation, where the errata record the disagreements that remain;
  - genomic (`g.`) and structural HGVS.

  See the [design contract](design/duckvep.md) and the [roadmap](https://github.com/RGenomicsETL/DuckVEP/issues).

## Speed

Recorded benchmark results use their stated hosts, inputs, output contracts and thread counts; they describe those runs.

- **SQL annotation:** a one-core compact-output run on GIAB HG002 measured **`r n(round(pub$variants_per_second))` alleles per second** across `r n(pub$variants)` model-addressable alleles, with `r n(pub$regulation_features)` regulatory and motif features resident. The measurement ran through `FROM query(duckvep_annotate_sql(...))`; see the [throughput method and receipts](benchmarks/duckvep_throughput.md).
- **Comparison with vep-rs:** three paired GIAB HG002 runs pinned to one E-core measured median process times of **58.09 s for DuckVEP and 73.26 s for vep-rs**, including disk-backed outputs on a shared host. Their common consequence tuples agree on 34,146,531 of 34,148,222 comparisons; executable VEP agrees with DuckVEP on 1,689 of the 1,691 disagreements. These are distinct output formats and measured conditions, not a general speed ranking. See the [plots, receipts and reproducible runner](benchmarks/benchmark_duckvep_vep_rs.md).
- **Model restore:** `duckvep_model_save` writes a loaded model snapshot and `duckvep_model_restore` maps it read-only. In an Ensembl 116 GRCh38 measurement, restore took about half a second, and processes restoring the same file share its mapped pages.
- **Memory limits:** native allocations are charged against an enforced budget; exceeding it raises a capacity error rather than truncating results. The [scale runner](docs/scale-runner.md) records three concurrent 5-million-allele gnomAD jobs on one 20-thread host.

## From R

`Rduckvep` runs the same native annotation builders on your DuckDB connection. Builder functions return SQL, and `rduckvep_annotate()` executes it and returns a data frame. Annotation queries can read relations on that connection, including TEMP tables and uncommitted rows.

```r
con <- Rduckvep::rduckvep_connect()
# After loading a model named "demo" and creating "demo_events":
sql <- Rduckvep::rduckvep_annotate_sql(con, "demo_events", "demo", hgvs = TRUE)
results <- Rduckvep::rduckvep_annotate(con, "demo_events", "demo", hgvs = TRUE)
DBI::dbDisconnect(con, shutdown = TRUE)
```

## With the rest of the stack

DuckVEP needs no other extension, but it is meant to sit in a query next to them. With [DuckHTS](https://github.com/RGenomicsETL/duckhts) reading the VCF and [DuckClinVarbitration](https://github.com/RGenomicsETL/DuckClinVarbitration) supplying arbitrated ClinVar decisions, annotation and clinical evidence meet in one query. This one isn't evaluated here, because it needs a full Ensembl model and both extensions:

```sql
-- One row per ALT allele, with the model's region index (DuckHTS reads the VCF)
CREATE TABLE alleles AS
SELECT row_number() OVER (ORDER BY r.seq_region, v.POS, alt_allele)::UBIGINT AS event_index,
       r.seq_region, v.POS::UBIGINT AS position, v.REF AS reference, alt_allele AS alternate,
       NULL::UBIGINT AS end_position, NULL::VARCHAR AS structural_type,
       NULL::VARCHAR AS copy_change, NULL::UINTEGER AS mate_seq_region,
       NULL::UBIGINT AS mate_position
FROM read_bcf('cohort.vcf.gz') v CROSS JOIN unnest(v.ALT) AS t(alt_allele)
JOIN grch38_regions r   -- Ensembl names contigs 1..22, X, Y, MT; UCSC-style VCFs say chr1, chrM
  ON r.seq_region_name = CASE WHEN v.CHROM IN ('chrM', 'M') THEN 'MT'
                              ELSE regexp_replace(v.CHROM, '^chr', '') END;

-- High-impact consequences next to their arbitrated ClinVar decision
SELECT r.seq_region_name AS contig, e.position, e.reference, e.alternate,
       a.consequence, a.protein_hgvs, d.policy_classification, d.gold_stars
FROM query(duckvep_annotate_sql('alleles', 'human_116_grch38', struct_pack(rich := true, hgvs := true))) a
JOIN alleles e USING (event_index)
JOIN grch38_regions r USING (seq_region)
LEFT JOIN clinvar_vcf v                              -- DuckClinVarbitration
  ON v.assembly = 'GRCh38' AND v.contig = r.seq_region_name AND v.position = e.position
 AND v.reference = e.reference AND v.alternate = e.alternate
LEFT JOIN clinvar_policy_allele_decisions d USING (allele_id)
WHERE a.impact = 'HIGH';
```

## Install and build

```sh
git clone --recurse-submodules https://github.com/RGenomicsETL/DuckVEP
cd DuckVEP
make configure release test_release
duckdb -unsigned -c "LOAD 'build/release/duckvep.duckdb_extension'"
```

DuckHTS releases through 1.5.2 register legacy DuckVEP functions under the same SQL names. Loading one alongside DuckVEP can split model loading and annotation across separate model registries, producing `unknown model name`. Use DuckHTS newer than 1.5.2 with DuckVEP.

The extension uses the stable DuckDB C API (tested on DuckDB 1.5 and the 2.0 pre-release) and links its own htslib (for indexed reference FASTA) and cgranges. From R, [Rduckvep](https://rgenomicsetl.github.io/DuckVEP/Rduckvep/) builds the same sources offline and adds connection, model and haplotype helpers.

## Documentation

- [Function reference](docs/functions.md): every public SQL function, its signature, options, columns and a checked example
- [Site](https://rgenomicsetl.github.io/DuckVEP/) and [R package reference](https://rgenomicsetl.github.io/DuckVEP/Rduckvep/)
- [Compatibility and errata](ERRATA.md): every known difference from VEP, classified, with witnesses
- [Design and implementation contract](design/duckvep.md): model build, the sweep, NMD, phased edits, structural events, HGVS, supplementary annotation
- [Haplotype contract `duckvep-coding`](design/duckvep_haplotype_contract.md) and its [scale qualification](benchmarks/data/haplotype_scale/README.md)
- Reports: [conformance](benchmarks/duckvep_conformance.md), [throughput](benchmarks/duckvep_throughput.md), [vep-rs comparison](benchmarks/benchmark_duckvep_vep_rs.md), [haplotypes](benchmarks/duckvep_haplotypes.md), [scale runner](docs/scale-runner.md)
- [Corpus workflow](design/duckvep_corpus_workflow.md) and [conformance harness](test/duckvep/conformance/README.md)

DuckVEP was developed within [DuckHTS](https://github.com/RGenomicsETL/duckhts) and extracted with its Git history.

## License

GPL-2.0-or-later; see [LICENSE](LICENSE). Vendored htslib and cgranges keep their own licences. Ensembl data and VEP are © EMBL-EBI under their own terms.
