---
output:
  github_document:
    pandoc_args: ["--wrap=preserve"]
---

<!-- README.md is generated from README.Rmd using duckknit. Please edit that file. -->

```{r setup, include=FALSE}
library(duckknit)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::knit_engines$set(sql = knitr::knit_engines$get("duckdb"))
knitr::knit_hooks$set(document = function(x) gsub("(?m)^``` duckdb", "```sql", x, perl = TRUE))
cli <- Sys.getenv("DUCKDB_CLI")
stopifnot(nzchar(cli), file.exists(cli))
options(duckknit.duckdb = cli)
local({
  ns <- asNamespace("duckknit")
  proc <- processx::process$new(cli, args = c("-unsigned", ":memory:"),
    stdin = "|", stdout = "|", stderr = "|", cleanup = TRUE, cleanup_tree = TRUE)
  unlockBinding(".sessions", ns)
  ns$.sessions[["readme"]] <- proc
  lockBinding(".sessions", ns)
  unlockBinding(".state", ns)
  ns$.state$last_session <- "readme"
  lockBinding(".state", ns)
  extension <- normalizePath("build/release/duckvep.duckdb_extension", mustWork = TRUE)
  load <- duckknit:::duckknit_exec(proc, sprintf("LOAD '%s';", extension))
  if (nzchar(trimws(load$stderr))) stop(load$stderr)
  fixture <- paste(readLines("test/data/duckvep/readme.sql"), collapse = "\n")
  setup <- duckknit:::duckknit_exec(proc, fixture)
  if (nzchar(trimws(setup$stderr))) stop(setup$stderr)
})
source("scripts/readme-evidence.R")
evidence <- readme_evidence()
n <- function(x) format(x, big.mark = ",", scientific = FALSE, trim = TRUE)
conf <- evidence$conformance
hgvs <- evidence$hgvs
fast <- evidence$fastvep
tp <- evidence$throughput
```

# DuckVEP

[![Extension CI](https://github.com/RGenomicsETL/DuckVEP/actions/workflows/MainDistributionPipeline.yml/badge.svg)](https://github.com/RGenomicsETL/DuckVEP/actions/workflows/MainDistributionPipeline.yml)
[![Site](https://github.com/RGenomicsETL/DuckVEP/actions/workflows/pages.yml/badge.svg)](https://rgenomicsetl.github.io/DuckVEP/)

**A variant consequence engine for DuckDB, with Ensembl VEP semantics and a ledger of every place it differs.**

Variant effect prediction usually means a Perl pipeline that reads a VCF and writes a text file. DuckVEP makes it a relational operator. An Ensembl release is compiled once into an immutable, shared transcript model; sorted alleles then stream through a native C sweep that emits typed rows: Sequence Ontology consequences, impact, cDNA/CDS/protein positions, amino acids, NMD prediction and HGVS. Those rows join with anything DuckDB can read (a VCF through [DuckHTS](https://github.com/RGenomicsETL/duckhts), Parquet, DuckLake, ClinVar, population frequencies) without a plugin system or a new file format.

The semantics are not a reinterpretation of the SO rules. They are defined by executable Ensembl VEP: every claim is measured against a pinned release of the real tool (today VEP `r sub("^VEP ", "", conf$oracle[[1L]])`, with its matching core and variation databases), every result that differs is kept in the denominator, and the places where DuckVEP deliberately behaves differently are written down. Moving to a new VEP release means re-running the ledger, not re-arguing the rules.

## How close to VEP

Transcript and feature consequence pairs compared with executable VEP, newest tested run per corpus:

```{r conformance, echo=FALSE}
shown <- conf
shown$pairs <- n(shown$pairs)
shown$exact <- n(shown$exact)
shown$unresolved <- n(shown$unresolved)
shown$differences <- n(shown$differences)
knitr::kable(shown, col.names = c("Corpus", "Assembly", "Oracle", "Pairs", "Exact", "Unresolved", "Different"),
  align = c("l", "l", "l", "r", "r", "r", "r"))
```

That is **`r n(sum(conf$exact))` of `r n(sum(conf$pairs))`** pairs identical to VEP, across human GRCh38 and GRCh37 and *Plasmodium falciparum* (with its own genetic codes), small variants, exact structural events, paired breakends, and regulatory and motif features. The release gate is deterministic: a single discordant, missing, extra or unresolved pair fails the run. HGVS is held to the same standard: `r n(hgvs$exact[hgvs$field == "HGVSc"])` HGVSc and `r n(hgvs$exact[hgvs$field == "HGVSp"])` HGVSp transcript pairs match VEP `--hgvs` string for string (`r n(hgvs$clinvar_pairs[[1L]])` of them from ClinVar chromosome 21), with `r n(sum(hgvs$discordant))` discordant.

Methods, per-term and per-impact tables, generated state-exploration campaigns and statistical caveats are in the [conformance report](benchmarks/duckvep_conformance.md). Ledger revisions predate the extraction from DuckHTS and are resolved through the [commit map](design/duckhts-commit-map.txt).

## Where it differs, and why

Compatibility is claimed only where it is measured, and differences are recorded rather than hidden. The [compatibility and errata record](ERRATA.md) keeps three verdicts apart (an observed VEP convention DuckVEP must follow, a DuckVEP gap, and a potential upstream erratum), each with its witness; the main cases:

- **Published release annotations are not the oracle.** Ensembl's release VCFs disagree with executable VEP at some sites (in release 116, `X/Y:276322 G>A` is published as `intergenic_variant`, while the executable emits three `5_prime_UTR_variant` rows per chromosome). DuckVEP follows the executable; the [PAR witnesses](test/duckvep/conformance/README.md) pin both.
- **Breakends are evaluated one event at a time.** VEP 116's buffered breakend path uses a chromosome-blind interval tree and can drop valid transcript pairs in multi-chromosome batches. The oracle runs with `--buffer_size 1` to isolate that; DuckVEP has no such batching effect ([paired-breakend differential](benchmarks/duckvep_conformance.md#paired-breakend-differential)).
- **Imprecise structural variants use the nominal span**, exactly as VEP's registered predicates do, while `CIPOS`/`CIEND` stay on the row as metadata instead of being dropped.
- **No silent size limits.** VEP skips structural events above `--max_sv_size` (5 kb by default); DuckVEP annotates every exact span, and the oracle is run with a 10 Mb limit so the comparison covers them.
- **Unknown is a value.** When a result cannot be computed (missing reference sequence, reference mismatch, an unsupported symbolic allele) the row carries `duckvep_status` and a reason instead of a guess.
- **Outside the claim:** untested species and releases, phased multi-record haplotypes against an executable oracle, compound (cis) HGVS presentation, where the errata record the retained disagreements, and genomic (`g.`) and structural HGVS. See the [design contract](design/duckvep.md) and the [roadmap](https://github.com/RGenomicsETL/DuckVEP/issues).

## Speed

On the full GIAB HG002 GRCh38 benchmark (`r n(evidence$giab_alt_alleles)` ALT alleles, VCF in, uncompressed table out, including DuckVEP's explicit coordinate sort), DuckVEP finished **`r sprintf("%.2f", fast$ratio[[1L]])`×** faster than FastVEP on one core (`r sprintf("%.0f", fast$duckvep[[1L]])` s against `r sprintf("%.0f", fast$fastvep[[1L]])` s) and **`r sprintf("%.2f", fast$ratio[[2L]])`×** on four, on a compact output contract. Complete-field comparisons, including CSQ with HGVS, and every caveat are in the [FastVEP benchmark](benchmarks/benchmark_duckvep_fastvep.md). The annotation kernel alone sustains `r n(round(tp$variants_per_second))` variants per second on one core with rich output ([throughput report](benchmarks/duckvep_throughput.md)).

## Try it

A model is loaded from ordinary relations: regions, transcripts with their CDS and flanking sequence, and exons. In production these come from an Ensembl core database through `duckvep_ensembl_regions()` and `duckvep_ensembl_transcripts()`; here a [one-transcript fixture](test/data/duckvep/readme.sql) stands in:

```{sql model, eval=TRUE}
SELECT loaded FROM duckvep_model_load('demo',
  'SELECT * FROM readme_regions ORDER BY seq_region',
  'SELECT * FROM readme_transcripts ORDER BY seq_region, transcript_start',
  'SELECT * FROM readme_exons ORDER BY transcript_index, exon_start');
```

Alleles are rows too. Seven variants across the transcript, from the 5′ UTR to the intron:

```{sql annotate, eval=TRUE}
CREATE TABLE demo_events AS
SELECT row_number() OVER (ORDER BY position, alternate)::UBIGINT AS event_index,
       1::UINTEGER AS seq_region, position::UBIGINT AS position, reference, alternate,
       NULL::UBIGINT AS end_position, NULL::VARCHAR AS structural_type,
       NULL::VARCHAR AS copy_change, NULL::UINTEGER AS mate_seq_region,
       NULL::UBIGINT AS mate_position
FROM (VALUES (110, 'C', 'T'), (124, 'T', 'C'), (125, 'A', 'G'), (130, 'C', 'CA'),
             (134, 'C', 'A'), (151, 'G', 'A'), (175, 'A', 'G')) v(position, reference, alternate);

SELECT e.position, e.reference || '>' || e.alternate AS change,
       a.consequence, a.impact, a.transcript_hgvs, a.protein_hgvs, a.nmd_prediction
FROM query(duckvep_annotate_sql('demo_events', 'demo', struct_pack(rich := true, hgvs := true))) a
JOIN demo_events e USING (event_index)
ORDER BY e.position;
```

The frameshift and the stop-gain escape nonsense-mediated decay because they fall in the first CDS positions (`nmd_escape_early_cds`). The empty HGVS fields are not failures to parse: this model was loaded without a reference FASTA, and those notations need flanking sequence to shift and render, so DuckVEP says so instead of guessing:

```{sql unresolved, eval=TRUE}
SELECT e.position, a.transcript_hgvs_status, a.transcript_hgvs_reason
FROM query(duckvep_annotate_sql('demo_events', 'demo', struct_pack(hgvs := true))) a
JOIN demo_events e USING (event_index)
WHERE a.transcript_hgvs IS NULL
ORDER BY e.position;
```

Phased genotypes become whole-haplotype consequences, again as a relation:

```{sql haplotypes, eval=TRUE}
CREATE TABLE demo_calls AS
SELECT event_index, seq_region, position, reference, alternate,
       1::UINTEGER alt_index, 0::UINTEGER transcript_index, 0::UBIGINT sample_index,
       [1, 0]::INTEGER[] alleles, [false, true]::BOOLEAN[] phase_before, NULL::BIGINT phase_set
FROM demo_events WHERE position IN (124, 125);

SELECT transcript_index, carrier_count, sequence_status, length(contributors) AS contributors
FROM duckvep_haplotypes('SELECT * FROM demo_calls', 'demo');
```

## With the rest of the stack

DuckVEP needs no other extension, but it is built to sit in a query next to them. With [DuckHTS](https://github.com/RGenomicsETL/duckhts) reading the VCF and [DuckClinVarbitration](https://github.com/RGenomicsETL/DuckClinVarbitration) supplying arbitrated ClinVar decisions, annotation and clinical evidence meet in one query (not evaluated here: it needs a full Ensembl model and both extensions):

```sql
-- One row per ALT allele, with the model's region index (DuckHTS reads the VCF)
CREATE TABLE alleles AS
SELECT row_number() OVER (ORDER BY r.seq_region, v.POS, alt_allele)::UBIGINT AS event_index,
       r.seq_region, v.POS::UBIGINT AS position, v.REF AS reference, alt_allele AS alternate,
       NULL::UBIGINT AS end_position, NULL::VARCHAR AS structural_type,
       NULL::VARCHAR AS copy_change, NULL::UINTEGER AS mate_seq_region,
       NULL::UBIGINT AS mate_position
FROM read_bcf('cohort.vcf.gz') v CROSS JOIN unnest(v.ALT) AS t(alt_allele)
JOIN grch38_regions r ON r.seq_region_name = v.CHROM;

-- High-impact consequences next to their arbitrated ClinVar decision
SELECT r.seq_region_name AS contig, e.position, e.reference, e.alternate,
       a.consequence, a.protein_hgvs, d.policy_classification, d.gold_stars
FROM query(duckvep_annotate_sql('alleles', 'human_116_grch38', struct_pack(rich := true, hgvs := true))) a
JOIN alleles e USING (event_index)
JOIN grch38_regions r USING (seq_region)
LEFT JOIN clinvar_vcf v                              -- DuckClinVarbitration
  ON v.assembly = 'GRCh38' AND v.contig = r.seq_region_name AND v.position = e.position
 AND v.reference = e.reference AND v.alternate = e.alternate
LEFT JOIN clinvar_policy_allele_decisions d USING (allele_id)
WHERE a.impact = 'HIGH';
```

## Install and build

```sh
git clone --recurse-submodules https://github.com/RGenomicsETL/DuckVEP
cd DuckVEP
make configure release test_release
duckdb -unsigned -c "LOAD 'build/release/duckvep.duckdb_extension'"
```

The extension uses the stable DuckDB C API (tested on DuckDB 1.5 and the 2.0 pre-release) and links its own htslib (for indexed reference FASTA) and cgranges. From R, [Rduckvep](https://rgenomicsetl.github.io/DuckVEP/Rduckvep/) builds the same sources offline and provides connection, model and haplotype helpers.

## Documentation

- [Site](https://rgenomicsetl.github.io/DuckVEP/) and [R package reference](https://rgenomicsetl.github.io/DuckVEP/Rduckvep/)
- [Compatibility and errata](ERRATA.md): every known difference from VEP, classified, with witnesses
- [Design and implementation contract](design/duckvep.md): model build, the sweep, NMD, phased edits, structural events, HGVS, supplementary annotation
- Reports: [conformance](benchmarks/duckvep_conformance.md), [throughput](benchmarks/duckvep_throughput.md), [FastVEP comparison](benchmarks/benchmark_duckvep_fastvep.md), [haplotypes](benchmarks/duckvep_haplotypes.md)
- [Corpus workflow](design/duckvep_corpus_workflow.md) and [conformance harness](test/duckvep/conformance/README.md)

DuckVEP was developed inside [DuckHTS](https://github.com/RGenomicsETL/duckhts) and extracted with its history.

## License

GPL-2.0-or-later; see [LICENSE](LICENSE). Vendored htslib and cgranges keep their own licences. Ensembl data and VEP are © EMBL-EBI under their own terms.
