---
title: "DuckVEP throughput"
output:
  github_document:
    html_preview: false
---

<!-- duckvep_throughput.md is generated from duckvep_throughput.Rmd. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
  dev = "ragg_png", dpi = 144, fig.width = 9, fig.height = 4.8,
  fig.path = "figures/throughput-", fig.align = "center")
root <- normalizePath(Sys.getenv("DUCKVEP_REPO_ROOT", getwd()), mustWork = TRUE)
library(ggplot2)
history <- read.csv(file.path(root, "benchmarks/data/duckvep_throughput.csv"),
  stringsAsFactors = FALSE, check.names = FALSE,
  colClasses = c(source_revision = "character", checksum_value = "character",
    fingerprint_xor_hash = "character", fingerprint_sum_hash = "character"))

surface_revision <- "660a4ed9a6710846d4e92d14c0630df3cb129053"
surface <- subset(history, source_revision == surface_revision &
  workload == "ensembl116_grch38_giab_hg002_v4_2_1_full_literal_public_relation" &
    output_mode %in% c("compact", "rich", "hgvs", "rich_hgvs") &
    threads %in% c(1L, 4L) & input_partitions == 1L &
    transcript_distance == 5000 & regulation_features == 1383580)
stopifnot(nrow(surface) == 8L,
  !anyDuplicated(surface[c("output_mode", "threads")]),
  all(surface$passes == 5L), all(surface$warmup_variants == 100000L),
  length(unique(surface$variants)) == 1L,
  length(unique(surface$annotated_rows)) == 1L,
  all(vapply(surface[c("model_database_sha256", "variants_database_sha256",
    "corpus_source_sha256")], function(x) length(unique(x)) == 1L, logical(1L))))
for (pair in split(surface, surface$output_mode)) {
  stopifnot(setequal(pair$threads, c(1L, 4L)),
    length(unique(pair$checksum_kind)) == 1L,
    length(unique(pair$checksum_value)) == 1L)
  if (all(pair$fingerprint_schema_version == 2L, na.rm = FALSE)) {
    stopifnot(length(unique(pair$fingerprint_xor_hash)) == 1L,
      length(unique(pair$fingerprint_sum_hash)) == 1L)
  }
}

core_labels <- c(
  ensembl116_grch38_clinvar_20260706_full_literal_regulation = "ClinVar",
  ensembl116_grch38_giab_hg002_v4_2_1_full_literal_regulation = "HG002")
core_revision <- "03803bd8dff47f0d41584e7fb7a79d42c9183f04"
core <- subset(history, source_revision == core_revision & workload %in% names(core_labels) &
  output_mode %in% c("compact", "rich", "hgvs") & threads == 1L &
  input_partitions == 1L & transcript_distance == 5000)
identity_columns <- c("extension_sha256", "model_database_sha256",
  "model_logical_sha256", "model_source_manifest_sha256", "model_reference_sha256",
  "model_region_ordinal_sha256", "variants_database_sha256", "corpus_source_sha256")
stopifnot(nrow(core) == 6L, !anyDuplicated(core[c("workload", "output_mode")]),
  all(core$regulation_features == 1383580),
  all(core$extension_build_binding == "htslib_distclean_make_release"),
  all(vapply(core[identity_columns], function(x) all(nchar(x) == 64L), logical(1L))),
  all(core$model_source_version == "116"), all(core$model_assembly == "GRCh38"),
  all(core$variants == core$corpus_supported_alt_count),
  all(core$fingerprint_schema_version == 2L),
  all(nzchar(core$fingerprint_xor_hash)), all(nzchar(core$fingerprint_sum_hash)),
  length(unique(core$extension_sha256)) == 1L,
  length(unique(core$model_database_sha256)) == 1L,
  length(unique(core$model_logical_sha256)) == 1L)
for (corpus in split(core, core$workload)) {
  stopifnot(setequal(corpus$output_mode, c("compact", "rich", "hgvs")),
    length(unique(corpus$variants)) == 1L,
    length(unique(corpus$annotated_rows)) == 1L)
}

dense <- read.csv(file.path(root, "benchmarks/data/duckvep_dense_threading.csv"),
  stringsAsFactors = FALSE, colClasses = c(source_revision = "character",
    xor_hash = "character", sum_hash = "character"))
stopifnot(nrow(dense) == 8L, length(unique(dense$source_revision)) == 1L,
  !anyDuplicated(dense[c("transcript_distance", "threads")]))
for (pair in split(dense, dense$transcript_distance)) {
  stopifnot(setequal(pair$threads, c(1L, 4L)),
    length(unique(pair$variants)) == 1L,
    length(unique(pair$annotated_rows)) == 1L,
    length(unique(pair$xor_hash)) == 1L, length(unique(pair$sum_hash)) == 1L,
    length(unique(pair$model_sha256)) == 1L,
    length(unique(pair$region_ordinal_sha256)) == 1L,
    length(unique(pair$variants_database_sha256)) == 1L)
}

human <- read.csv(file.path(root, "benchmarks/data/duckvep_human_annotation.csv"),
  stringsAsFactors = FALSE, colClasses = c(source_revision = "character"))
human_fingerprint <- read.csv(file.path(root,
  "benchmarks/data/duckvep_human_annotation_fingerprint.csv"), colClasses = "character")
stopifnot(nrow(human) == 13L, length(unique(human$input_alleles)) == 1L,
  length(unique(human$threads)) == 1L, nrow(human_fingerprint) == 2L,
  all(vapply(human_fingerprint[c("rows", "xor64", "sum64")],
    function(x) length(unique(x)) == 1L, logical(1L))))
composition_stages <- c("model_load", "vcf_decode_canonicalize_sort",
  "DuckVEP_rich_HGVS_regulatory_parallel_stream_4GB",
  "compose_all_providers_parallel_stream_4GB")
composition <- human[match(composition_stages, human$stage), , drop = FALSE]
stopifnot(!anyNA(composition$stage))
composition$query <- factor(c("Model load", "VCF decode + sort",
  "Annotation + HGVS + write", "Annotation + all joins + write"),
  levels = rev(c("Model load", "VCF decode + sort",
    "Annotation + HGVS + write", "Annotation + all joins + write")))
composition$label <- sprintf("%.2f s", composition$seconds)

interval <- read.csv(file.path(root,
  "benchmarks/data/duckvep_human_interval_validation.csv"), stringsAsFactors = FALSE)
stopifnot(nrow(interval) == 3L, all(interval$mismatched_event_rows == 0L),
  all(vapply(interval[c("matched_events", "overlap_pairs", "stable_ids", "so_terms")],
    function(x) length(unique(x)) == 1L, logical(1L))))
interval$method_label <- factor(c("Chromosome hash + residuals", "Packed RegionKey IEJoin",
  "cgranges bulk index"), levels = rev(c("Chromosome hash + residuals",
    "Packed RegionKey IEJoin", "cgranges bulk index")))

mode_labels <- c(compact = "Compact", rich = "Rich", hgvs = "HGVS",
  rich_hgvs = "Rich + HGVS")
mode_colors <- c(compact = "#0072B2", rich = "#009E73", hgvs = "#D55E00",
  rich_hgvs = "#8E5EA2")
thread_colors <- c(`1 core` = "#6B7280", `4 cores` = "#087E8B")
plot_theme <- theme_minimal(base_size = 14) + theme(
  plot.background = element_rect(fill = "white", color = NA),
  panel.grid.minor = element_blank(), panel.grid.major.y = element_blank(),
  plot.title = element_text(face = "bold", size = 18),
  plot.title.position = "plot",
  plot.subtitle = element_text(color = "#4B5563", size = 11),
  axis.title = element_text(size = 12), legend.position = "top",
  legend.title = element_blank(), strip.text = element_text(face = "bold"),
  plot.margin = margin(12, 32, 12, 12))

surface$output <- factor(mode_labels[surface$output_mode], levels = rev(mode_labels))
surface$workers <- factor(paste(surface$threads, "core", sep = ""),
  levels = c("1core", "4core"), labels = c("1 core", "4 cores"))
surface$label <- sprintf("%.2f s", surface$median_seconds)
core$output <- factor(mode_labels[core$output_mode], levels = rev(mode_labels[1:3]))
core$corpus <- paste0(unname(core_labels[core$workload]), " | ",
  sprintf("%.2fM alleles, %.1f rows/allele", core$variants / 1e6,
    core$annotated_rows / core$variants))
core$rate <- core$variants_per_second / 1e6
core$rate_label <- sprintf("%.2fM / s", core$rate)
dense$workers <- factor(ifelse(dense$threads == 1L, "1 core", "4 cores"),
  levels = names(thread_colors))
dense$halo_kb <- dense$transcript_distance / 1000
```

**Throughput depends on the output you request and the rows each allele produces.**
The charts separate pinned resident annotation from dense-region threading and
whole-genome composition. [Methods, measurement scope and reproduction](duckvep_throughput_evidence.md)
cover the details; [CSV receipts](data/duckvep_throughput.csv) retain every recorded run.

## Public relation API cost and scaling

The same HG002 relation emits `r format(unique(surface$annotated_rows), big.mark=",", scientific=FALSE)`
rows at either worker count. Compact, rich, HGVS and fused rich+HGVS each pass their
cross-thread output checks.

```{r public-cost, fig.alt="HG002 wall time for four output surfaces, comparing one and four pinned cores; four cores reduce each median to about one third.", fig.height=5.2}
ggplot(surface, aes(median_seconds, output, fill = workers)) +
  geom_col(position = position_dodge(0.78), width = 0.64) +
  geom_errorbar(aes(xmin = min_seconds, xmax = max_seconds),
    position = position_dodge(0.78), width = 0.15, orientation = "y") +
  geom_text(aes(label = label), position = position_dodge(0.78),
    hjust = -0.15, size = 3.8) +
  scale_fill_manual(values = thread_colors) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(title = "More output costs time; four cores share the work",
    subtitle = paste0(unique(surface$run_date), " | ", substr(surface_revision, 1, 8),
      " | five passes | whiskers: observed min–max"),
    x = "Wall time (seconds; lower is better)", y = NULL) + plot_theme
```

The public relation dispatcher distributes ordered input across workers. One core
is pinned to CPU 2; four cores to 2,4,6,8. The receipt records CPU-set contention.

## Full-corpus core VEP model

These one-core measurements use the complete Ensembl 116 GRCh38 transcript,
regulatory and motif model. Timers include result production and aggregation;
loading and file ingestion are outside the measurement.

```{r core-rates, fig.alt="One-core input-allele throughput for ClinVar and HG002, faceted by corpus, with compact, rich and cumulative-HGVS output shown separately.", fig.height=6.6}
ggplot(core, aes(rate, output, fill = output_mode)) +
  geom_col(width = 0.62) +
  geom_errorbar(aes(xmin = variants / max_seconds / 1e6,
    xmax = variants / min_seconds / 1e6), width = 0.15, orientation = "y") +
  geom_text(aes(label = rate_label), hjust = -0.15, size = 3.8) +
  facet_wrap(~corpus, ncol = 1) +
  scale_fill_manual(values = mode_colors, guide = "none") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.22))) +
  labs(title = "Output detail changes the input rate",
    subtitle = paste0(unique(core$run_date), " | ", substr(core_revision, 1, 8),
      " | pinned one-core medians | whiskers: observed min–max"),
    x = "Input alleles / second (millions; higher is better)", y = NULL) + plot_theme
```

Rows per allele differ between corpora. An input rate alone does not describe the
amount of output produced.

## Dense-region threading and halo

This fixture has 517,097 ClinVar alleles in 318 annotation-dense tiles. Each halo
pairs one core with four ordered partitions; matching fingerprints verify the
same output rows.

```{r halo-throughput, fig.alt="Dense-region input throughput at 0, 5, 10 and 50 kilobase transcript halos; four pinned cores outperform one at every tested halo."}
ggplot(dense, aes(halo_kb, variants / median_seconds / 1000,
    color = workers, group = workers)) +
  geom_line(linewidth = 0.9) + geom_point(size = 3) +
  scale_color_manual(values = thread_colors) +
  scale_x_continuous(breaks = c(0, 5, 10, 50)) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.08))) +
  labs(title = "A wider halo does more work per allele",
    subtitle = paste0(substr(unique(dense$source_revision), 1, 8), " | compact output | five-pass medians"),
    x = "Transcript halo (kb)", y = "Input alleles / second (thousands)") + plot_theme
```

```{r halo-fanout, fig.alt="Dense-region output grows from 18.2 million rows at zero halo to 88.8 million at 50 kilobases, from the same 517097 input alleles.", fig.height=4.4}
fanout <- dense[dense$threads == 1L, , drop = FALSE]
ggplot(fanout, aes(halo_kb, annotated_rows / 1e6)) +
  geom_line(color = "#087E8B", linewidth = 0.9) +
  geom_point(color = "#087E8B", size = 3) +
  geom_text(aes(label = sprintf("%.1fM", annotated_rows / 1e6)),
    vjust = -0.8, size = 4) +
  scale_x_continuous(breaks = c(0, 5, 10, 50)) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.14))) +
  labs(title = "The halo expands the output, not the input",
    subtitle = "One and four cores emit identical row multisets at each distance",
    x = "Transcript halo (kb)", y = "Output rows (millions)") + plot_theme
```

```{r dense-memory, fig.alt="Peak process RSS for one and four pinned cores stays near 5.2 GiB across all four tested transcript halos.", fig.height=4.4}
ggplot(dense, aes(halo_kb, maximum_rss_kib / 1024^2,
    color = workers, group = workers)) +
  geom_line(linewidth = 0.9) + geom_point(size = 3) +
  scale_color_manual(values = thread_colors) +
  scale_x_continuous(breaks = c(0, 5, 10, 50)) +
  scale_y_continuous(limits = c(0, 6), breaks = 0:6) +
  annotate("text", x = 25, y = 2.7,
    label = sprintf("Both configurations: %.2f–%.2f GiB RSS",
      min(dense$maximum_rss_kib) / 1024^2, max(dense$maximum_rss_kib) / 1024^2),
    size = 5, color = "#4B5563") +
  labs(title = "Four workers keep the same process footprint",
    subtitle = "GNU time peak process RSS; includes model load and DuckDB, not just C workspace",
    x = "Transcript halo (kb)", y = "Peak process RSS (GiB)") + plot_theme
```

Measurements cover these halos and four partitions on the annotation-dense panel.
RSS differences are process measurements, not allocation attribution.

## Real whole-genome composition run

A separate DeepVariant HG002 WGS run writes complete rich+HGVS output joined to
ClinVar, ClinvArbitration, AlphaMissense, gene constraint and Ensembl regulation.
It uses four writers, warm page cache and no CPU pinning.

```{r composition-queries, fig.alt="Standalone whole-genome statements: model load and VCF preparation, annotation plus Parquet write, and annotation with all resource joins plus write.", fig.height=4.8}
ggplot(composition, aes(seconds, query)) +
  geom_col(fill = "#087E8B", width = 0.6) +
  geom_text(aes(label = label), hjust = -0.15, size = 4) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(title = "From a VCF to a fully joined Parquet relation",
    subtitle = "Warm-page-cache statements | four writers | full queries overlap: times are not additive",
    x = "Statement time (seconds; lower is better)", y = NULL) + plot_theme
```

The all-provider query writes `r format(composition$result_rows[4], big.mark=",", scientific=FALSE)`
rows in `r sprintf("%.2f", composition$seconds[4])` seconds, with
`r sprintf("%.2f", composition$max_rss_kb[4] / 1024^2)` GiB peak RSS. These are
integration measurements, separate from the pinned resident rates.

## Interval join plans

All three implementations return the same 745,252 overlap pairs and match the
same 414,813 alleles. The physical join plan controls the work.

```{r interval-time, fig.alt="Equivalent interval joins take 83.96 seconds with a chromosome hash join and residual predicates, 0.77 seconds with packed RegionKey IEJoin, and 1.144 seconds with cgranges.", fig.height=4.5}
ggplot(interval, aes(seconds, method_label)) +
  geom_col(fill = "#087E8B", width = 0.6) +
  geom_text(aes(label = sprintf("%.3f s", seconds)), hjust = -0.15, size = 4) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.16))) +
  labs(title = "Equivalent rows, very different join costs",
    subtitle = "EXPLAIN identifies the physical plan; result equivalence is checked per allele",
    x = "Statement time (seconds; lower is better)", y = NULL) + plot_theme
```

```{r interval-memory, fig.alt="Peak RSS for packed RegionKey IEJoin is 1.52 GiB versus 0.79 GiB for cgranges; the chromosome-hash receipt has no memory measurement.", fig.height=3.6}
measured_interval <- interval[!is.na(interval$max_rss_kb), , drop = FALSE]
ggplot(measured_interval, aes(max_rss_kb / 1024^2, droplevels(method_label))) +
  geom_col(fill = "#087E8B", width = 0.56) +
  geom_text(aes(label = sprintf("%.2f GiB", max_rss_kb / 1024^2)),
    hjust = -0.15, size = 4) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(title = "cgranges uses less memory; IEJoin finishes sooner",
    subtitle = "No RSS measurement is recorded for the chromosome-hash query",
    x = "Peak process RSS (GiB; lower is better)", y = NULL) + plot_theme
```

## Evidence and reproduction

[Methods and scope](duckvep_throughput_evidence.md) define the timers, output
surfaces and equality checks. CSVs retain full provenance and raw measurements:
[resident annotation](data/duckvep_throughput.csv),
[dense threading](data/duckvep_dense_threading.csv),
[whole-genome statements](data/duckvep_human_annotation.csv), and
[interval equivalence](data/duckvep_human_interval_validation.csv).

Render the plots and this report from the repository root:

```sh
Rscript benchmarks/scripts/render_benchmarks.R throughput
```
